---
title: "Democratization, Personal Wealth of Politicians and Voting Behavior"
author: | 
  | Bas Machielsen
  | EHS 2022
institute: "Utrecht University"
email: "a.h.machielsen@uu.nl"
date: \today
output: 
  beamer_presentation:
    includes:
      in_header: preamble.tex
classoption: "aspectratio=169"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(stargazer)
library(tidyverse)
library(ivreg)
library(knitr)

```

# Motivation

## Motivation

- Between 1850 to 1920, a wave of democratization and liberalization swept over Western Europe, bringing about universal suffrage and an expansion of government. 

- The double transition from autocracy to parliamentary democracy, and from passive government to engagement in social spending, has been widely studied in various disciplines. 
  - Threat of revolution (Acemoglu and Robinson, 2000)
  - Electoral expedience (Lizzeri and Perisco, 2004)
  - Electoral competition (Llavador and Oxoby, 2005)
  
- The conception of politicians' self-interest in these studies revolves around safeguarding political power. 

- However, politicians might also care about a more superficial form of self-interest:  their personal wealth (Ferraz and Finan, 2009; Tahoun and Van Lent,2019).

## This study

- I use the setting of the Netherlands to analyze politicians' voting behavior in parliament on all suffrage extensions and major fiscal legislation between 1872-1921
  - Inheritance taxation (\textit{Successiewet})
  - Income taxation (\textit{Inkomstenbelasting})

- I employ probate inventories collected from archival sources to identify the relationship between politicians' personal wealth and their voting behavior. 

- Using data on the portfolio composition of politicians' wealth, I estimate politicians' net wealth at the time of voting, and relate this to the voting outcome.

- I provide instrumental variable (IV) estimates of personal wealth on the propensity to vote for reforms.
  - Politician's wealth is instrumented by parental wealth and by an indicator whether the politician’s father was also politically active.

## Results - Overview

- Major finding: richer politicians are more likely to vote against fiscal legislation than _ceteris paribus_ poorer politicians

  - Instrumental variable analyses suggest there is a direct effect of fiscal legislation on personal future wealth. 

- The effects are absent in suffrage extensions
  
- Consistent with the historiography, ideology and party are major determinants of voting behavior

  - But the effect of personal wealth is strong enough to materially influence the voting outcome
  
- Paradigm-shifting laws are the primary drivers for the effect
  
  - Consistent with the effect being more pronounced when expected costs of acceptance are higher

# Background


## Legislation and Politicians

- __Suffrage extensions:__ Initially, very few enfranchised: based on tax burden. 
  - Failed attempt at expansion in 1872, extensions in 1887, 1896 and 1917 (universal male suffrage). 
  - Main causes of disagreement: specific numbers and precise criteria. Not a debate dominated by abstract principles. 
  - Large effect on the electorate, but likely no effect for politicians personally.

- __Fiscal legislation:__ 1850-1870: government size was very limited, no structural increase in government expenditures. 
  - In 1878, a law project passed which introduced inheritance taxation for lineal descendants.
  - In 1893, first income taxation introduced. Precedents for successive increases in rates. 
  - Potentially large effect on politicians' personal finances. 
  
## Parliamentary Wealth over Time

```{r, out.width="0.7\\linewidth", include=TRUE, fig.align="center", fig.cap=c("Parliamentary Wealth over Time"), echo=FALSE}
knitr::include_graphics("../figures/wealth_parl.pdf")
```

## Personal Impact on Politicians

```{r, out.width="0.7\\linewidth", include=TRUE, fig.align="center", fig.cap=c("Tax Rates on Income (Left) and Wealth (Right)"), echo=FALSE}
knitr::include_graphics("../figures/tax_rates.pdf")
```


# Method

## Empirical Model


- I firstly pool all laws, and model $V = \text{Pr}(p_i = 1)$ as a function of a politician's wealth and party, augmented by controls:

$$
V_{i, j} = \alpha + \beta \cdot \text{ihs } \text{Wealth}_{i,j} + \delta \cdot \text{Party}_{i} + \gamma \cdot \text{LawDum}_{j} + \eta \cdot \text{Controls}_{i,j} + \varepsilon_{i,j}
$$

- In other specifications, I separately estimate regressions for suffrage extensions and fiscal legislation.

- I use the following recursive relationship to estimate a politician's wealth at the time of voting on law $k$ as a function of their (deflated) wealth at death: 

$$
\text{Wealth}_{i,t+1} = \sum_{J} \text{AssetShare}_{i,j,t} \cdot \text{AssetReturn}_{i,j,[t,t+1]}
$$

## Two Ways to Address Endogeneity

- Even after correcting for differential wealth returns, politicians' wealth could be endogenously determined.
  - Particular voting behavior might be rewarded by interest groups (Fisman et al., 2014, Tahoun and van Lent, 2019)
  - I test whether there is a difference in the personal wealth - voting behavior relationship between politicians who died shortly after having vote and those who did not. 

- There might still be many reasons why this approach does not isolate the effect of personal wealth on voting behavior. 
  - E.g. politicians’ consumption and investment behavior might be correlated with their voting behavior: politicians who vote against might have consumed more of their income, reducing the observed correlation between personal wealth and  voting  behavior.  
  - More generally, unobserved politician fixed effects might be correlated with wealth. 
  
- I use instrumental variable estimation to address this. 
  - Two unrelated instruments: _Father Politician_ and _Expected Inheritance_

# Results

## Descriptive Statistics

\input{../tables/descriptivestats_dissent.tex}

## OLS Estimates

- OLS Estimates of Wealth on the Propensity to Vote for Suffrage and Fiscal Legislation
  - Results show small but significant relationship

\begin{table}[!h]

\centering
\resizebox{\paperheight}{!}{
\begin{tabular}[t]{lccccc}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{3}{c}{Pooled} & \multicolumn{1}{c}{Suffrage} & \multicolumn{1}{c}{Fiscal} \\
\cmidrule(l{3pt}r{3pt}){2-4} \cmidrule(l{3pt}r{3pt}){5-5} \cmidrule(l{3pt}r{3pt}){6-6}
  & Model 1 & Model 2 & Model 3 & Model 4 & Model 5\\
\midrule
Personal Wealth &  & \num{-0.007}*** &  & \num{-0.008} & \num{-0.009}*\\
 &  & (\num{0.003}) &  & (\num{0.005}) & (\num{0.005})\\
Personal Wealth x Fiscal &  &  & \num{-0.007}** &  & \\
 &  &  & (\num{0.004}) &  \vphantom{1} & \\
Personal Wealth x Suffrage &  &  & \num{-0.007}* &  & \\
 &  &  & (\num{0.004}) &  & \\
Liberal & \num{0.385}*** & \num{0.433}*** & \num{0.433}*** & \num{0.383}*** & \num{0.373}***\\
 & (\num{0.027}) & (\num{0.033}) & (\num{0.033}) & (\num{0.071}) & (\num{0.062})\\
Socialist & \num{0.523}*** & \num{0.602}*** & \num{0.602}*** & \num{0.568}*** & \num{0.281}***\\
 & (\num{0.034}) & (\num{0.047}) & (\num{0.047}) & (\num{0.113}) & (\num{0.080})\\
\midrule
Law Fixed Effects & Yes & Yes & Yes & Yes & Yes\\
N & \num{1028} & \num{675} & \num{675} & \num{249} & \num{270}\\
Adj. R2 & \num{0.35} & \num{0.36} & \num{0.36} & \num{0.29} & \num{0.43}\\
\bottomrule
\multicolumn{6}{l}{\rule{0pt}{1em}Heteroskedasticity-robust standard errors in parentheses. Results for lower house voting outcomes.}\\
\multicolumn{6}{l}{\rule{0pt}{1em}Control variables omitted from the table.}\\
\multicolumn{6}{l}{\rule{0pt}{1em}* p $<$ 0.1, ** p $<$ 0.05, *** p $<$ 0.01}\\
\end{tabular}}
\end{table}

## Alternative Explanations

- The effect of personal wealth on the likelihood to vote in favor of fiscal legislation seems to be robust to the inclusion of many controls

- But politicians who voted against could have been rewarded by interest groups
  - Wealth at death is higher because of voting profile $\rightarrow$ coefficient might be _overestimated_

- Test whether there is a difference in the relationship between wealth and voting behavior for politicians who died early and those who did not
  - The group who died shortly after voting is unlikely to have had opportunities to amass rents
  - If there is a substantial difference, reverse causality could play a large role
  
## Alternative Explanations: Died Shortly After Vote

- The coefficient on personal wealth is significant in both analyses, as well as in the pooled case. 


\begin{table}
\centering
\resizebox{\paperheight}{!}{
\begin{tabular}[t]{lcccccc}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Pooled} & \multicolumn{2}{c}{Suffrage} & \multicolumn{2}{c}{Fiscal} \\
\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5} \cmidrule(l{3pt}r{3pt}){6-7}
  & Model 1 & Model 2 & Model 3 & Model 4 & Model 5 & Model 6\\
\midrule
Personal Wealth & \num{-0.008}* & \num{-0.010}** & \num{-0.012}* & \num{-0.013}* & \num{-0.009} & \num{-0.016}*\\
 & (\num{0.004}) & (\num{0.004}) & (\num{0.007}) & (\num{0.007}) & (\num{0.006}) & (\num{0.008})\\
Died W 2 Yrs & \num{0.028} & \num{0.033} & \num{0.032} & \num{0.023} & \num{0.031} & \num{-0.036}\\
 & (\num{0.063}) & (\num{0.069}) & (\num{0.109}) & (\num{0.107}) & (\num{0.095}) & (\num{0.107})\\
Personal Wealth x Died W 2 Yrs & \num{0.002} & \num{0.006} & \num{0.011} & \num{0.013} & \num{0.002} & \num{0.009}\\
 & (\num{0.006}) & (\num{0.006}) & (\num{0.009}) & (\num{0.009}) & (\num{0.008}) & (\num{0.009})\\
Liberal & \num{0.457}*** & \num{0.376}*** & \num{0.365}*** & \num{0.361}*** & \num{0.389}*** & \num{0.368}***\\
 & (\num{0.034}) & (\num{0.045}) & (\num{0.070}) & (\num{0.072}) & (\num{0.058}) & (\num{0.061})\\
Socialist & \num{0.496}*** & \num{0.381}*** & \num{0.612}*** & \num{0.600}*** & \num{0.327}*** & \num{0.277}***\\
 & (\num{0.043}) & (\num{0.060}) & (\num{0.106}) & (\num{0.111}) & (\num{0.072}) & (\num{0.086})\\
\midrule
Law Fixed Effects & Yes & Yes & Yes & Yes & Yes & Yes\\
N & \num{629} & \num{571} & \num{257} & \num{249} & \num{314} & \num{275}\\
Adj. R2 & \num{0.33} & \num{0.36} & \num{0.30} & \num{0.31} & \num{0.41} & \num{0.42}\\
\bottomrule
\multicolumn{7}{l}{\rule{0pt}{1em}Heteroskedasticity-robust standard errors in parentheses. Results for lower house voting outcomes.}\\
\multicolumn{7}{l}{\rule{0pt}{1em}Control variables omitted from the table.}\\
\multicolumn{7}{l}{\rule{0pt}{1em}* p $<$ 0.1, ** p $<$ 0.05, *** p $<$ 0.01}\\
\end{tabular}}
\end{table}

## IV Estimates

- There are still various reasons why this coefficient might not reflect the true relationship between personal wealth and voting behavior.

- Politicians’ consumption and investment behavior might be correlated with their voting behavior: politicians who vote against might have consumed more of their income, reducing the observed correlation between personal wealth and voting behavior.  

- More generally, unobserved politician fixed effects might be correlated with wealth (omitted variable bias)
  
- I use instrumental variable estimation to address this. 
  - Two unrelated instruments: _Father Politician_ and _Expected Inheritance_
  
## IV Estimates: Suffrage Extensions

- Father Politician is a good predictor for politicians' wealth. 
- No precise relationship between personal wealth and propensity to vote in favor of suffrage extension.

\begin{table}[!h]
\centering
\resizebox{0.8 \textwidth}{!}{
\begin{tabular}[t]{lcccccc}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{1}{c}{Personal Wealth} & \multicolumn{1}{c}{Vote} & \multicolumn{1}{c}{Personal Wealth} & \multicolumn{1}{c}{Vote} & \multicolumn{1}{c}{Personal Wealth} & \multicolumn{1}{c}{Vote} \\
\cmidrule(l{3pt}r{3pt}){2-2} \cmidrule(l{3pt}r{3pt}){3-3} \cmidrule(l{3pt}r{3pt}){4-4} \cmidrule(l{3pt}r{3pt}){5-5} \cmidrule(l{3pt}r{3pt}){6-6} \cmidrule(l{3pt}r{3pt}){7-7}
  & Model 1 & Model 2 & Model 3 & Model 4 & Model 5 & Model 6\\
\midrule
Father Politician & \num{2.125}*** &  & \num{1.826}** &  & \num{1.329}* & \\
 & (\num{0.676}) &  & (\num{0.788}) &  & (\num{0.793}) & \\
Personal Wealth &  & \num{-0.028} &  & \num{-0.046} &  & \num{-0.085}\\
 &  & (\num{0.030}) &  & (\num{0.042}) &  & (\num{0.076})\\
Liberal & \num{1.376}** & \num{0.446}*** & \num{0.930} & \num{0.394}*** & \num{1.134}* & \num{0.400}***\\
 & (\num{0.667}) & (\num{0.069}) & (\num{0.667}) & (\num{0.086}) & (\num{0.683}) & (\num{0.120})\\
Socialist & \num{0.768} & \num{0.500}*** & \num{3.782}*** & \num{0.699}*** & \num{4.397}*** & \num{0.847}**\\
 & (\num{1.800}) & (\num{0.107}) & (\num{0.992}) & (\num{0.183}) & (\num{1.375}) & (\num{0.338})\\
\midrule
Law Fixed Effects & Yes & Yes & Yes & Yes & Yes & Yes\\
Kleibergen-Paap F Stat. &  & 9.5 &  & 6.43 &  & 3.77\\
N & \num{236} & \num{236} & \num{213} & \num{213} & \num{180} & \num{180}\\
Adj. R2 & \num{0.05} & \num{0.23} & \num{0.07} & \num{0.13} & \num{0.17} & \num{-0.16}\\
\bottomrule
\multicolumn{7}{l}{\rule{0pt}{1em}Heteroskedasticity-robust standard errors in parentheses. Results for lower house voting outcomes.}\\
\multicolumn{7}{l}{\rule{0pt}{1em}Controls omitted from the table.}\\
\multicolumn{7}{l}{\rule{0pt}{1em}* p $<$ 0.1, ** p $<$ 0.05, *** p $<$ 0.01}\\
\end{tabular}}
\end{table}

## IV Estimates: Fiscal Legislation

- Father Politician is a good predictor for politicians' wealth. 
- Significant and negative relationship between personal wealth and propensity to vote in favor of suffrage extension.
  - About 2-3 times the effect size of the OLS analyses.

\begin{table}[!h]
\centering
\resizebox{0.8 \textwidth}{!}{
\begin{tabular}[t]{lcccccc}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{1}{c}{Personal Wealth} & \multicolumn{1}{c}{Vote} & \multicolumn{1}{c}{Personal Wealth} & \multicolumn{1}{c}{Vote} & \multicolumn{1}{c}{Personal Wealth} & \multicolumn{1}{c}{Vote} \\
\cmidrule(l{3pt}r{3pt}){2-2} \cmidrule(l{3pt}r{3pt}){3-3} \cmidrule(l{3pt}r{3pt}){4-4} \cmidrule(l{3pt}r{3pt}){5-5} \cmidrule(l{3pt}r{3pt}){6-6} \cmidrule(l{3pt}r{3pt}){7-7}
  & Model 1 & Model 2 & Model 3 & Model 4 & Model 5 & Model 6\\
\midrule
Father Politician & \num{2.965}*** &  & \num{2.424}*** &  & \num{2.156}*** & \\
 & (\num{0.497}) &  & (\num{0.515}) &  & (\num{0.570}) & \\
Personal Wealth &  & \num{-0.041}** &  & \num{-0.049}** &  & \num{-0.062}**\\
 &  & (\num{0.018}) &  & (\num{0.023}) &  & (\num{0.031})\\
Liberal & \num{1.349}** & \num{0.522}*** & \num{1.971}*** & \num{0.454}*** & \num{1.567}** & \num{0.448}***\\
 & (\num{0.592}) & (\num{0.052}) & (\num{0.745}) & (\num{0.078}) & (\num{0.766}) & (\num{0.086})\\
Socialist & \num{1.977}* & \num{0.534}*** & \num{3.823}** & \num{0.450}*** & \num{3.957}** & \num{0.447}***\\
 & (\num{1.125}) & (\num{0.067}) & (\num{1.547}) & (\num{0.120}) & (\num{1.687}) & (\num{0.150})\\
\midrule
Law Fixed Effects & Yes & Yes & Yes & Yes & Yes & Yes\\
Kleibergen-Paap F Stat. &  & 20.58 &  & 13.41 &  & 12.36\\
Selection Ratio &  &  &  & 20.88 &  & 1.04\\
N & \num{346} & \num{346} & \num{312} & \num{312} & \num{240} & \num{240}\\
Adj. R2 & \num{0.11} & \num{0.22} & \num{0.11} & \num{0.22} & \num{0.13} & \num{0.18}\\
\bottomrule
\multicolumn{7}{l}{\rule{0pt}{1em}Heteroskedasticity-robust standard errors in parentheses. Results for lower house voting outcomes.}\\
\multicolumn{7}{l}{\rule{0pt}{1em}* p $<$ 0.1, ** p $<$ 0.05, *** p $<$ 0.01}\\
\end{tabular}}
\end{table}

## IV Estimates: Fiscal Legislation - Expected Inheritance

- Relationship persists when using a completely unrelated (and also plausible) instrument. 

\begin{table}[!h]
\centering
\resizebox{0.8 \textwidth}{!}{
\begin{tabular}[t]{lcccccc}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{1}{c}{Personal Wealth} & \multicolumn{1}{c}{Vote} & \multicolumn{1}{c}{Personal Wealth} & \multicolumn{1}{c}{Vote} & \multicolumn{1}{c}{Personal Wealth} & \multicolumn{1}{c}{Vote} \\
\cmidrule(l{3pt}r{3pt}){2-2} \cmidrule(l{3pt}r{3pt}){3-3} \cmidrule(l{3pt}r{3pt}){4-4} \cmidrule(l{3pt}r{3pt}){5-5} \cmidrule(l{3pt}r{3pt}){6-6} \cmidrule(l{3pt}r{3pt}){7-7}
  & Model 1 & Model 2 & Model 3 & Model 4 & Model 5 & Model 6\\
\midrule
Expected Inheritance & \num{0.262}*** &  & \num{0.237}** &  & \num{0.235}*** & \\
 & (\num{0.088}) &  & (\num{0.111}) &  & (\num{0.084}) & \\
Personal Wealth &  & \num{-0.031}* &  & \num{-0.033}* &  & \num{-0.041}**\\
 &  & (\num{0.017}) &  & (\num{0.017}) &  & (\num{0.020})\\
Liberal & \num{1.573} & \num{0.544}*** & \num{2.389}* & \num{0.545}*** & \num{0.754} & \num{0.401}***\\
 & (\num{1.042}) & (\num{0.073}) & (\num{1.318}) & (\num{0.124}) & (\num{0.899}) & (\num{0.141})\\
Socialist & \num{0.082} & \num{0.511}*** & \num{1.440} & \num{0.478}*** & \num{0.029} & \num{0.349}**\\
 & (\num{1.553}) & (\num{0.087}) & (\num{1.993}) & (\num{0.144}) & (\num{1.623}) & (\num{0.163})\\
\midrule
Law Fixed Effects & Yes & Yes & Yes & Yes & Yes & Yes\\
Kleibergen-Paap F Stat. &  & 2.33 &  & 2.1 &  & 2.41\\
Selection Ratio &  &  &  & 0.63 &  & 0.55\\
N & \num{171} & \num{171} & \num{152} & \num{152} & \num{108} & \num{108}\\
Adj. R2 & \num{0.07} & \num{0.27} & \num{0.08} & \num{0.28} & \num{0.14} & \num{0.29}\\
\bottomrule
\multicolumn{7}{l}{\rule{0pt}{1em}Heteroskedasticity-robust standard errors in parentheses. Results for lower house voting outcomes.}\\
\multicolumn{7}{l}{\rule{0pt}{1em}* p $<$ 0.1, ** p $<$ 0.05, *** p $<$ 0.01}\\
\end{tabular}}
\end{table}

## Impact on Probability of Acceptance

```{r, out.width="0.9\\linewidth", include=TRUE, fig.align="center", fig.cap=c("Wealth and Probability of Acceptance"), echo=FALSE}
knitr::include_graphics("../figures/interpretation.pdf")
```

# Conclusion

## Conclusion

- I find a significant and persistent negative effect of politicians' wealth on the tendency to vote in favor of tax increases
  - Dissenting politicians tend to prioritize their finances

- The effect is economically significant: poorer parliaments would have approved rejected laws sooner, and richer parliaments would have rejected laws increasing taxation that have been accepted. 

- Clear relationship between personal wealth and voting behavior is absent in the case of suffrage extension

- Implications for democratization literature
  - Direct evidence that the personal profile of politicians influences government size
  - Partial explanation for why government size increased slowly over the nineteenth century (cf. Lindert, 2004) is the domination of politics by wealthy elites. 
  - Exogenous negative wealth shocks facilitated expansion.

# Robustness Checks

## Robustness Checks - Logit Results

\begin{table}[!h]
\caption{\label{tab:logit_suffrage_fiscal}Logit Analysis of Suffrage Extension and Fiscal Legislation}
\centering
\resizebox{0.8 \paperheight}{!}{
\begin{tabular}[t]{lcccccc}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{3}{c}{Suffrage} & \multicolumn{3}{c}{Fiscal} \\
\cmidrule(l{3pt}r{3pt}){2-4} \cmidrule(l{3pt}r{3pt}){5-7}
  & Model 1 & Model 2 & Model 3 & Model 4 & Model 5 & Model 6\\
\midrule
Personal Wealth & \num{-0.039} & \num{-0.042} & \num{-0.038} & \num{-0.049}* & \num{-0.063}* & \num{-0.076}*\\
 & (\num{0.029}) & (\num{0.030}) & (\num{0.032}) & (\num{0.029}) & (\num{0.034}) & (\num{0.039})\\
Number of Strikes &  & \num{0.060} & \num{0.046} &  & \num{0.007} & \num{-0.081}\\
 &  & (\num{0.064}) & (\num{0.061}) &  & (\num{0.031}) & (\num{0.131})\\
Vote Share &  & \num{-0.484} & \num{-0.615} &  & \num{0.006} & \num{0.786}\\
 &  & (\num{0.771}) & (\num{0.785}) &  & (\num{0.882}) & (\num{1.022})\\
Turnout &  & \num{0.075} & \num{-0.337} &  & \num{0.161} & \num{-0.516}\\
 &  & (\num{0.853}) & (\num{0.919}) &  & (\num{1.099}) & (\num{1.285})\\
Margin to Nearest Competitor &  & \num{-0.779} & \num{-0.804} &  & \num{-0.356} & \num{-0.404}\\
 &  & (\num{1.009}) & (\num{1.030}) &  & (\num{0.968}) & (\num{1.093})\\
Tenure &  & \num{-0.019} & \num{-0.018} &  & \num{-0.005} & \num{-0.035}\\
 &  & (\num{0.020}) & (\num{0.020}) &  & (\num{0.021}) & (\num{0.023})\\
Share Catholic &  &  & \num{-0.249} &  &  & \num{-3.130}***\\
 &  &  & (\num{0.643}) &  &  & (\num{0.831})\\
Share Tax Liable in District &  &  & \num{5.445} &  &  & \num{30.544}\\
 &  &  & (\num{16.118}) &  &  & (\num{20.464})\\
\midrule
Party Fixed Effects & Yes & Yes & Yes & Yes & Yes & Yes\\
Law Fixed Effects & Yes & Yes & Yes & Yes & Yes & Yes\\
N & \num{282} & \num{260} & \num{249} & \num{342} & \num{315} & \num{270}\\
$R^2$ & \num{0.01} & \num{0.03} & \num{0.03} & \num{0.01} & \num{0.01} & \num{0.10}\\
Max. $R^2$ & \num{0.58} & \num{0.59} & \num{0.58} & \num{0.48} & \num{0.49} & \num{0.50}\\
\bottomrule
\multicolumn{7}{l}{\rule{0pt}{1em}The dependent variable, Vote, is defined as 1 if the politician is in favor of the reform, 0 otherwise.}\\
\multicolumn{7}{l}{\rule{0pt}{1em}The reference political allegiance is confessional.}\\
\multicolumn{7}{l}{\rule{0pt}{1em}Standard errors in parentheses. Results for lower house voting outcomes.}\\
\multicolumn{7}{l}{\rule{0pt}{1em}* p $<$ 0.1, ** p $<$ 0.05, *** p $<$ 0.01}\\
\end{tabular}}
\end{table}


## Robustness Checks - Wealth at Death

\begin{table}[!h]
\caption{\label{tab:fisc_nw0_ols_iv}IV Analysis of Fiscal Legislation - Robustness Check}
\centering
\resizebox{\paperheight}{!}{
\begin{tabular}[t]{lcccccc}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{3}{c}{OLS} & \multicolumn{3}{c}{IV} \\
\cmidrule(l{3pt}r{3pt}){2-4} \cmidrule(l{3pt}r{3pt}){5-7}
  & Model 1 & Model 2 & Model 3 & Model 4 & Model 5 & Model 6\\
\midrule
Personal Wealth & \num{-0.007}* & \num{-0.006} & \num{-0.009}* & \num{-0.037}** & \num{-0.046}** & \num{-0.049}**\\
 & (\num{0.004}) & (\num{0.004}) & (\num{0.004}) & (\num{0.016}) & (\num{0.021}) & (\num{0.022})\\
Liberal & \num{0.482}*** & \num{0.384}*** & \num{0.373}*** & \num{0.510}*** & \num{0.456}*** & \num{0.420}***\\
 & (\num{0.043}) & (\num{0.054}) & (\num{0.057}) & (\num{0.051}) & (\num{0.074}) & (\num{0.070})\\
Socialist & \num{0.518}*** & \num{0.337}*** & \num{0.332}*** & \num{0.558}*** & \num{0.482}*** & \num{0.482}***\\
 & (\num{0.077}) & (\num{0.103}) & (\num{0.119}) & (\num{0.089}) & (\num{0.144}) & (\num{0.160})\\
\midrule
Law Fixed Effects & Yes & Yes & Yes & Yes & Yes & Yes\\
Kleibergen-Paap F Stat. &  &  &  & 20.46 & 13.01 & 13.5\\
N & \num{347} & \num{311} & \num{272} & \num{351} & \num{316} & \num{277}\\
Adj. R2 & \num{0.36} & \num{0.41} & \num{0.43} & \num{0.22} & \num{0.20} & \num{0.24}\\
\bottomrule
\multicolumn{7}{l}{\rule{0pt}{1em}Heteroskedasticity-robust standard errors in parentheses. Results for lower house voting outcomes.}\\
\multicolumn{7}{l}{\rule{0pt}{1em}* p $<$ 0.1, ** p $<$ 0.05, *** p $<$ 0.01}\\
\end{tabular}}
\end{table}

## Robustness Checks - Log Wealth

\begin{table}[!h]

\caption{\label{tab:ivresults_fisc_log}IV Estimates of Wealth on the Propensity to Vote for Fiscal Reforms}
\centering
\resizebox{\paperheight}{!}{
\begin{tabular}[t]{lcccccc}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{1}{c}{Personal Wealth} & \multicolumn{1}{c}{Vote} & \multicolumn{1}{c}{Personal Wealth} & \multicolumn{1}{c}{Vote} & \multicolumn{1}{c}{Personal Wealth} & \multicolumn{1}{c}{Vote} \\
\cmidrule(l{3pt}r{3pt}){2-2} \cmidrule(l{3pt}r{3pt}){3-3} \cmidrule(l{3pt}r{3pt}){4-4} \cmidrule(l{3pt}r{3pt}){5-5} \cmidrule(l{3pt}r{3pt}){6-6} \cmidrule(l{3pt}r{3pt}){7-7}
  & Model 1 & Model 2 & Model 3 & Model 4 & Model 5 & Model 6\\
\midrule
Father Politician & \num{1.368}*** &  & \num{1.287}*** &  & \num{1.567}*** & \\
 & (\num{0.177}) &  & (\num{0.189}) &  & (\num{0.221}) & \\
Personal Wealth &  & \num{-0.084}** &  & \num{-0.093}** &  & \num{-0.091}**\\
 &  & (\num{0.038}) &  & (\num{0.043}) &  & (\num{0.040})\\
Liberal & \num{0.406}** & \num{0.520}*** & \num{0.377} & \num{0.384}*** & \num{0.348} & \num{0.380}***\\
 & (\num{0.189}) & (\num{0.050}) & (\num{0.247}) & (\num{0.064}) & (\num{0.280}) & (\num{0.073})\\
Socialist & \num{-0.679} & \num{0.423}*** & \num{-0.062} & \num{0.249}** & \num{1.462}** & \num{0.341}***\\
 & (\num{0.427}) & (\num{0.078}) & (\num{0.624}) & (\num{0.098}) & (\num{0.601}) & (\num{0.113})\\
\midrule
Law Fixed Effects & Yes & Yes & Yes & Yes & Yes & Yes\\
Kleibergen-Paap F Stat. &  & 43.07 &  & 33.97 &  & 42.59\\
N & \num{321} & \num{321} & \num{292} & \num{292} & \num{230} & \num{230}\\
Adj. R2 & \num{0.20} & \num{0.31} & \num{0.15} & \num{0.33} & \num{0.19} & \num{0.33}\\
\bottomrule
\multicolumn{7}{l}{\rule{0pt}{1em}Heteroskedasticity-robust standard errors in parentheses. Results for lower house voting outcomes.}\\
\multicolumn{7}{l}{\rule{0pt}{1em}* p $<$ 0.1, ** p $<$ 0.05, *** p $<$ 0.01}\\
\end{tabular}}
\end{table}

## Robustness Checks - Exclusion Restriction

- Threats to identification imply that instrument might in itself suffer from endogeneity bias.
  - Politicians whose fathers were politically active could share a latent encompassing ideology (e.g. a penchant for increasing the size of the government)
  - They could therefore be inclined to vote in favor of laws that expand government

- Robustness checks to make exclusion restriction plausible: government intervention
  - In this set of laws, it is unlikely that wealth affects voting behavior (No direct effect).
  - Any empirical correlation is then due to an (indirect) _latent political family-effect_. 
  - An empirical correlation would thus imply a violation of the exclusion restriction. 

## Robustness Checks - Exclusion Restriction

- No indication of direct effect of Political Family on voting behavior.
  - Implying the exclusion restriction for the other set of laws likely holds. 

\begin{table}[!h]
\centering
\resizebox{\paperheight}{!}{
\begin{tabular}[t]{lccccccc}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{5}{c}{OLS} & \multicolumn{2}{c}{IV} \\
\cmidrule(l{3pt}r{3pt}){2-6} \cmidrule(l{3pt}r{3pt}){7-8}
\multicolumn{1}{c}{ } & \multicolumn{1}{c}{-} & \multicolumn{1}{c}{Count} & \multicolumn{1}{c}{Dummy} & \multicolumn{1}{c}{Count} & \multicolumn{1}{c}{Dummy} & \multicolumn{1}{c}{-} & \multicolumn{1}{c}{Dummy} \\
\cmidrule(l{3pt}r{3pt}){2-2} \cmidrule(l{3pt}r{3pt}){3-3} \cmidrule(l{3pt}r{3pt}){4-4} \cmidrule(l{3pt}r{3pt}){5-5} \cmidrule(l{3pt}r{3pt}){6-6} \cmidrule(l{3pt}r{3pt}){7-7} \cmidrule(l{3pt}r{3pt}){8-8}
  & Model 1 & Model 2 & Model 3 & Model 4 & Model 5 & Model 6 & Model 7\\
\midrule
Personal Wealth & \num{0.000} & \num{0.001} & \num{0.001} & \num{-0.001} & \num{-0.001} & \num{-0.039} & \num{0.002}\\
 & (\num{0.002}) & (\num{0.003}) & (\num{0.003}) & (\num{0.003}) & (\num{0.003}) & (\num{0.143}) & (\num{0.014})\\
Political Family &  & \num{-0.007} & \num{-0.017} & \num{-0.009} & \num{-0.010} &  & \\
 &  & (\num{0.015}) & (\num{0.029}) & (\num{0.021}) & (\num{0.038}) &  & \\
Liberal & \num{0.432}*** & \num{0.430}*** & \num{0.431}*** & \num{0.444}*** & \num{0.445}*** & \num{0.531} & \num{0.434}***\\
 & (\num{0.031}) & (\num{0.031}) & (\num{0.031}) & (\num{0.047}) & (\num{0.046}) & (\num{0.383}) & (\num{0.060})\\
Socialist & \num{0.518}*** & \num{0.519}*** & \num{0.519}*** & \num{0.518}*** & \num{0.517}*** & \num{0.616} & \num{0.496}***\\
 & (\num{0.039}) & (\num{0.040}) & (\num{0.039}) & (\num{0.066}) & (\num{0.066}) & (\num{0.439}) & (\num{0.082})\\
\midrule
Law Fixed Effects & Yes & Yes & Yes & Yes & Yes & Yes & Yes\\
Kleibergen-Paap F Stat. &  &  &  &  &  & 0.4 & 27.95\\
N & \num{727} & \num{727} & \num{727} & \num{469} & \num{469} & \num{334} & \num{399}\\
Adj. R2 & \num{0.46} & \num{0.46} & \num{0.46} & \num{0.43} & \num{0.43} & \num{0.22} & \num{0.42}\\
\bottomrule
\multicolumn{8}{l}{\rule{0pt}{1em}Personal Wealth is instrumented by Father Politician (Model 6) and Political Family (Model 7).}\\
\multicolumn{8}{l}{\rule{0pt}{1em}Heteroskedasticity-robust standard errors in parentheses. Results for lower house voting outcomes.}\\
\multicolumn{8}{l}{\rule{0pt}{1em}* p $<$ 0.1, ** p $<$ 0.05, *** p $<$ 0.01}\\
\end{tabular}}
\end{table}

# Appendix

## Specification Endogeneity Test

- I use a dummy variable indicating whether a politician died within $x \in \{2,5\}$  years after having cast the vote on a particular law, and estimate the following model:

\begin{multline*}
V_{i, j} = \alpha + \beta_1 \cdot \text{ihs } \text{Wealth}_{i,j} + \beta_2 \cdot \text{Died within X years}_i + \\
\beta_3 \cdot \text{ihs } \text{Wealth x Died within X years }_{i,j} + \\
\beta_4 \cdot \text{Party}_i + \beta_5 \cdot \text{LawDum}_j + \gamma \cdot \text{Controls}_{i,j} + \varepsilon_{i,j}
\end{multline*}

<!--
## Descriptive Statistics

- Preliminary evidence of the importance of wealth


## Descriptive Statistics 

- Variation in the degree of unanimity per party

\input{../tables/descriptive_stats_per_party_lh.tex}

## Descriptive Statistics

- Variation is lower in upper house

\input{../tables/descriptive_stats_per_party_uh.tex}

# Results

## Results - Baseline model

- This estimate addresses the influence of wealth  _within-party_.
- Plausible magnitude: a 1% increase in wealth decreases the probability of voting in favor with 1.8 percentage point.

## Baseline Results - Full controls (TK)

- Effect keeps existing after controlling for religion, region, constituent interests, political position


## Baseline Results - Full controls (EK)

- In the Upper house, the results are less robust to the inclusion of controls
  - Though the coefficient borders on significance and the point estimate is negative


## Alternative Explanations

- The effect of personal wealth on the likelihood to vote in favor of fiscal legislation seems to be robust to many controls. 
- But: politicians who vote against could have been rewarded by interest groups 
  - Wealth at death higher because of voting profile: coefficient might be _overestimated_
  
- Analysis of politicians who died shortly after last vote
  - No opportunity to amass rents
  
- IV using parental wealth
  - Parental wealth is a good proxy for wealth at the time of voting
  - Parental wealth correlated with politicians' wealth, but no direct effect on voting behavior.

## Alternative Explanations: Died Shortly After Vote

- Dying shortly after leaving office leaves little time to accumulate wealth from interest groups
  

## Alternative Explanations: Parental Wealth

- Wealth of politicians' parents proxies for politicians' wealth before voting
    - No direct influence on politicians' voting behavior (other than through politicians' wealth)

## IV Estimates (Preliminary)

\label{iv}

# Conclusion

## Conclusion

- I find a significant and persistent effect of personal interests on voting behavior of politicians.

  - The counterfactual implies that if a politician's wealth would increase by 20%, the effects of personal interests subsume party or ideological allegiance. 
  - The effect seems to be persistent in the lower house. 

- Politicians prioritize not only their ideology, but also their finances
  - But only when faced with imminent and real prospects of losing/winning
  - No convincing "general equilibrium" effects
  
- Confessional politicians have less party discipline

- Future: more observations in IV regression
  
# Thank you for your attention!

# Appendix: Fiscal reform

## How far did fiscal reforms reach?

\label{fiscal}

- After about 1870, majority of government revenues came from __taxation__. 


## Fiscal Reforms

- The share of tax revenue that came from real estate and wealth remained low
  - Steadily rising after the introduction of the 1893 income tax
  - Fiscal reforms encompassed (mildly) _progressive_ taxes
  - Lagged effects


## Fiscal Reforms

- Variation in plausible impact on politicians' personal wealth:

- Politicians' wealth would have certainly have been impacted by the Successiewet:
  - Tariffs up to 10 % (1878)
  - The poorest (net wealth < 1,000 guilders) were exempted
  - The 1911 revision implied a sharp increase in tariffs for lineal descendants

- The income tax was progressive, but the maximum tariff implied liability of only 3,2% of earned income (Fritschy, 2005)

- The revision in 1914 implied a maximum marginal tariff of 5%. (Own calculations, IB1914)

# Appendix: Data

## Voting Records

\label{votingrecords}

- Voting behavior in various important votes that lead to the acceptance or rejection of various far-reaching laws throughout the period 1870-1920.

- _Handelingen_, [statengeneraaldigitaal.nl](http://www.statengeneraaldigitaal.nl)
  
  - Find the dates on which the final roll call vote on a law took place

  - Data contain a _verbatim_ transcription of debates leading up to a vote
  
  - List of names of politicians who voted in favor, and against

## Politicians' Wealth

\label{memories}

- I use the _Memories van Successie_: probate inventories were administered by the Dutch tax administration for the purpose of levying inheritance taxes 
    - Universal from 1877 onwards
    - Publicly available until 1927

- Detailing the inventory of all assets and debts held by the deceased. 

- Detailed description also allows us to research the effect of **asset composition** on likelihood to vote in favor of fiscal legislation. 

## Controls

- Data regarding politicians' careers, district representation, ideological and party affiliation, and social origin from the [Politiek Documentatie Centrum](https://www.pdc.nl/)

- Electoral controls: turnout, nearest competitor, newspaper recommendation, etc. (for lower house members) from [this repository](http://resources.huygens.knaw.nl/verkiezingentweedekamer) by Ron De Jong

- Data on the frequency and location of strikes from the [IISG Labour Conflicts database](https://collab.iisg.nl/web/labourconflicts/datafiles)

- Data on demographic and economic characteristics of districts from a [database administered by Onno Boonstra](https://datasets.iisg.amsterdam/dataverse/HDNG?q=&types=files&sort=dateSort&order=desc&page=1)
    - Aggregated from the municipality to the district level (if applicable)


# Appendix: Spatial descriptive statistics

## Descriptive Statistics

# Appendix: Method

## Empirical Challenges

- Reverse causality: 
    - Politicians' voting behavior might be rewarded by interest groups. 
    - Politicians who vote against laws
    - Risk to __overestimate__ the effect of personal wealth on voting behavior
    
- Instrumental variable: parental wealth
    - Assumed to be strongly correlated with politicians' wealth before voting
    - No direct influence on politicians' voting behavior (other than through politicians' wealth)
    
## Empirical Model

- Baseline model (politician $i$, roll-call vote $k$):

\[
V_{i, k} = \alpha + \beta_1 \cdot \text{Wealth}_{i,k} + \beta_2 \cdot \text{Wealth}_{i,k} \cdot \text{LawDum}_k + \gamma \cdot \text{Controls}_{i,k} + u_{i,k}
\]

- In other words, I pool laws and estimate a linear probability model

- Calculate wealth at the time of voting from wealth at death:
\[
\text{Wealth}_{i,t+1} = \sum_{J} \text{AssetShare}_{i,j,t} \cdot \text{AssetReturn}_{i,j,[t,t+1]}
\]

- Since I know wealth at death, and the portfolio composition, calculate the (average) real returns on asset class $j$ and estimate the wealth at the time of voting.

- Historical returns provided by Jorda et al. (2019)


# Appendix: Additional results

## Robustness Checks: Other Laws




## Robustness Checks: Social Redistribution

\label{otherresults}

## Robustness Checks: Suffrage Expansion


## Robustness Checks: Wealth Composition


# The End

-->

