---
title: "Personal Wealth and Voting Behavior of Politicians"
author: "Bas Machielsen"
institute: "Utrecht University"
email: "a.h.machielsen@uu.nl"
date: "1/29/2021"
output: 
  beamer_presentation:
    includes:
      in_header: preamble.tex
classoption: "aspectratio=169"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(stargazer)
library(tidyverse)

figures <- readRDS("../figures/figures_presentation.RDS")
map1878 <- readRDS("../figures/presentation_map_1878.RDS")
map1893 <- readRDS("../figures/presentation_map_1893.RDS")

descr <- readRDS("../figures/descr.RDS")

first_regs <- readRDS("../figures/first_regs.RDS")
second_order <- readRDS("../figures/second_order_regs.RDS")
```

# Motivation


## Motivation

- Many (Western) European countries underwent radical changes between roughly 1850 and 1920.

    - All of them experienced high economic growth and were rapidly industrializing.
    - Most of them transitioned to universal suffrage
    - Some of them established the groundwork of social security.

- Politicians can either obstruct or facilitate these changes

- What determines the behavior of politicians to do so?


## Motivation

- Common visions on what determines the voting behavior of politicians:
    
    - Electoral responsiveness: the principal-agent view of politics implies that well-monitoring politicians act in the interest of their principal, the electorate (Peltzman, 1982; Levitt, 1996)
    
    - Special cases of which are the threat of revolution (Aidt & Franck, 2019), or more broader constituent interests (Kalt & Zupan, 1984)
    
    - Interest groups (Fisman et al., 2004; Mian et al., 2010)
    
    - Ideology: politicians have an innate conviction or preference for a certain degree of redistribution.
    
- In this paper:

    - Does self-interest of politicians influence their voting behavior?
    
## Motivation

- I focus on several important laws that instigated __fiscal reforms__ in the Netherlands throughout the late 19th and early 20th centuries

- Excellent case to find out whether politicians prioritize their self-interest:

    - Fiscal reforms are very likely to affect politicians' personal finances. 
    - Various parliaments, laws with varying impact
    - Varying degree of consensus
    - Other high-profile laws to check robustness
    
- Implications for the role of politicians in (inclusive) economic development 

# Fiscal Reforms

## How far did fiscal reforms reach?

- After about 1870, majority of government revenues came from __taxation__. 

```{r echo=FALSE, out.width="250px", out.height="175px", fig.align='center'}
figures[[1]]
```

## Fiscal Reforms

- The share of tax revenue that came from real estate and wealth remained low
  - Steadily rising after the introduction of the 1893 income tax
  - Fiscal reforms encompassed (mildly) _progressive_ taxes
  - Lagged effects

```{r echo = FALSE, out.width="250px", out.height="150px", fig.align='center'}
figures[[2]]
```

## Fiscal Reforms

- Variation in plausible impact on politicians' personal wealth:

- Politicians' wealth would have certainly have been impacted by the Successiewet:
  - Tariffs up to 10 % (1878)
  - The poorest (net wealth < 1,000 guilders) were exempted
  - The 1911 implied a sharp increase in tariffs for lineal descendants

- The income tax was progressive, but the maximum tariff implied liability of only 3,2% of earned income (Fritschy, 2005)

- The revision in 1914 implied a maximum marginal tariff of 5%. (Own calculations, IB1914)

# Data

## Voting Records

- Voting behavior in various important votes that lead to the acceptance or rejection of various far-reaching laws throughout the period 1870-1920.

- _Handelingen_, [statengeneraaldigitaal.nl](http://www.statengeneraaldigitaal.nl)
  
  - Find the data on which the final roll call vote on a law took place

  - Data contain a _verbatim_ transcription of debates leading up to a vote
  
  - List of names of politicians who voted in favor, and against


## Politicians' Wealth

- I use the _Memories van Successie_: probate inventories were administered by the Dutch tax administration for the purpose of levying inheritance taxes 
    - Universal from 1877 onwards
    - Publicly available until 1927

- Detailing the inventory of all assets and debts held by the deceased. 

- Detailed description also allows us to research the effect of **asset composition** on likelihood to vote in favor of fiscal legislation. 

## Controls

- Data regarding politicians' careers, district representation, ideological and party affiliation, and social origin from the [Politiek Documentatie Centrum](https://www.pdc.nl/)

- Electoral controls: turnout, nearest competitor, newspaper recommendation, etc. (for lower house members) from [this repository](http://resources.huygens.knaw.nl/verkiezingentweedekamer) by Ron De Jong

- Data on the frequency and location of strikes from the [IISG Labour Conflicts database](https://collab.iisg.nl/web/labourconflicts/datafiles)

- Data on demographic and economic characteristics of districts from a [database administered by Onno Boonstra](https://datasets.iisg.amsterdam/dataverse/HDNG?q=&types=files&sort=dateSort&order=desc&page=1)
    - Aggregated from the municipality to the district level (if applicable)
    
## Descriptive Statistics

- Preliminary evidence of the importance of wealth

```{r echo = FALSE}
descr[[1]] %>%
  knitr::kable(caption = "Eerste Kamer", 
               digits = 0)

```

## Descriptive Statistics

```{r echo = FALSE}
descr[[2]] %>%
  knitr::kable(caption = "Tweede Kamer",
               digits = 0)
```
## Descriptive Statistics

- A correlation between region (religion) and voting behavior?

```{r echo = FALSE, out.width="300px", out.height="150px", fig.align='center'}
map1878
```

## Descriptive Statistics

```{r  echo = FALSE, out.width="300px", out.height="150px", fig.align='center'}
map1893
```

# Method

## Empirical Challenges

- Reverse causality: 
    - Politicians' voting behavior might be rewarded by interest groups. 
    - Politicians who vote against laws
    - Risk to __overestimate__ the effect of personal wealth on voting behavior
    
- Instrumental variable: parental wealth
    - Assumed to be strongly correlated with politicians' wealth before voting
    - No direct influence on politicians' voting behavior (other than through politicians' wealth)
    
## Empirical Model

- Baseline model (politician $i$, roll-call vote $k$):

\[
V_{i, k} = \alpha + \beta_1 \cdot \text{Wealth}_{i,k} + \beta_2 \cdot \text{Wealth}_{i,k} \cdot \text{LawDum}_k + \gamma \cdot \text{Controls}_{i,k} + u_{i,k}
\]

- In other words, I pool laws and estimate a linear probability model

- Calculate wealth at the time of voting from wealth at death:
\[
\text{Wealth}_{i,t+1} = \sum_{J} \text{AssetShare}_{i,j,t} \cdot \text{AssetReturn}_{i,j,[t,t+1]}
\]

- Since I know wealth at death, and the portfolio composition, calculate the (average) real returns on asset class $j$ and estimate the wealth at the time of voting.

- Historical returns provided by Jorda et al. (2019)


# Results

## Baseline Results

[Simple table, infavor vs. not and wealth]



## Baseline Results - Regression

- This estimate addresses the influence of wealth  _within-party_.
- Plausible magnitude: a 1% increase in wealth decreases the probability of voting in favor with 1.8 percentage point.

```{r echo = FALSE, results='asis'}
stargazer(first_regs,
          covariate.labels = c("Wealth"),
          dep.var.labels = "Vote",
          omit = c("law", "class"),
          omit.stat = c("adj.rsq", "ser","f"),
          add.lines = list(c("House", "Both", "Both", "TK", "EK"),
                           c("Controls", "None", "Party", "Party", "Party")
          ),
          notes.append = T,
          notes = c("Robust standard errors in parentheses"),
          header = F,
          font.size = "tiny",
          column.sep.width = "1pt",
          title = "Baseline regressions"
)

```

## Baseline Results - Full controls

- Effect keeps existing after controlling for religion, region, constituent interests

## IV Estimates (Preliminary)

- Control for the possible influence of interest groups

## Robustness: Died Shortly After Vote

- Shortly dying after voting leaves little time to accumulate wealth from interest groups

## Robustness Checks: Other Laws

[Effect on Electoral Expansion (0)]

[Effect on Social Intervention (0)]

## Robustness Checks: Wealth Composition

```{r echo = FALSE, results='asis'}
stargazer(second_order,
          covariate.labels = c("Real Estate", "Foreign Assets", "Shares"),
          dep.var.labels = "Vote",
          omit = c("law", "class"),
          omit.stat = c("adj.rsq", "ser","f"),
          add.lines = list(c("House", "Both", "Both", "Both", "Both", "Both", "Both"),
                           c("Controls", "Party", "Party+Law", 
                             "Party", "Party+Law",
                             "Party", "Party+Law")
          ),
          notes.append = T,
          notes = c("Robust standard errors in parentheses"),
          header = F,
          font.size = "tiny",
          column.sep.width = "1pt",
          title = "Second-order effects"
)

```

# Conclusion

## Conclusion

- I find a large and significant effect of personal interests on voting behavior of politicians.

- Consistent with the view that politicians are imperfectly monitored
  - They are imperfect agents of the principal
  
- Politicians prioritize not only their ideology, but also their finances
  - But only when faced with imminent and real prospects of losing/winning
  - No convincing "general equilibrium" effects
  
# Thank you for your attention!

