---
title: "Democratization, Personal Wealth of Politicians and Voting Behavior"
author: "Bas Machielsen"
institute: "Utrecht University"
email: "a.h.machielsen@uu.nl"
date: \today
output: 
  beamer_presentation:
    includes:
      in_header: preamble.tex
classoption: "aspectratio=169"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(stargazer)
library(tidyverse)
library(ivreg)
library(knitr)

```

# Motivation

## Motivation

- From about 1850 to 1920, a wave of democratization and liberalization swept over Western Europe, bringing about universal suffrage and an expansion of government. 

- The double transition from autocracy to parliamentary democracy, and from passive government to a welfare state, has been widely studied in various disciplines. 
  - Threat of revolution (Acemoglu and Robinson, 2000)
  - Electoral expedience (Lizzeri and Perisco, 2004)
  - Electoral competition (Llavador and Oxoby, 2005)
  
- The conception of politicians' self-interest in these studies revolves around safeguarding political power. 

- However, politicians might also care about a more superficial form of self-interest:  their personal wealth (Ferraz and Finan, 2009; Tahoun and Van Lent,2019).

## This study

- I use the setting of the Netherlands to analyze politicians' voting behavior in parliament on all suffrage extensions and tax hikes between 1872-1921
  - Inheritance taxation (\textit{Successiewet})
  - Income taxation (\textit{Inkomstenbelasting})

- I employ newly-collected probate inventories collected from various archival sources to identify the relationship between politicians' personal wealth and their voting behavior. 

- Using data on the portfolio composition of politicians' wealth, I estimate their net wealth at the time of voting, and relate this to the voting outcome.

- I provide instrumental variable (IV) estimates of personal wealth on the propensity to vote for reforms.
  - Politician's wealth is instrumental by parental wealth by an indicator whether the politician’s father was also politically active.
  - I undertake various efforts to make the exclusion restriction plausible.

## Results - Overview

- Major finding: richer politicians are more likely to vote against fiscal legislation than _ceteris paribus_ poorer politicians

  - Instrumental variable analyses suggest there is a direct effect of fiscal legislation on personal future wealth. 

  - The effects are robust to the inclusion of many control variables and alternative instruments, but they are absent in legislation _not directly affecting_ politicians' personal wealth (suffrage extensions)
  
- Consistent with the historiography, ideology and party are major determinants of voting behavior

  - But the effect of personal wealth is strong enough to materially influence the voting outcome
  
- Pioneering laws are the primary drivers for the effect
  
  - Consistent with politicians the effect being more pronounced when expected costs of acceptance are higher

# Background

## Suffrage Extensions

- From 1848 onward, government formation and legislative power were subjected to parliamentary control

- Initially, very few very enfranchised: based on tax burden (\textit{Censuskiesrecht}) (van der Kolk et al., 2018)

  - Failed attempt at expansion in 1872: questions of suffrage and income taxation were intertwined
  - In 1887, constitutional revisions granted suffrage to significantly more men, other criteria than taxation included. 
  - Followed by another failed attempt in 1892 and a significant expansion in 1896.

- Main causes of disagreement: specific numbers and precise criteria. Not a debate dominated by abstract principles. 

## Fiscal Legislation

- From the 1850’s onward, the government oversaw liberalization and harmonization in all sorts of domains (Knippenberg et al., 2000).

-  Nevertheless government size was still very limited, and while defense spending slightly decreased following more modest geopolitical ambitions (Van den Berg and Vis, 2013), government expenditures per capita did not see a structural increase (van Riel, 2018).

- This started to change from the 1870's onward: 
  - In 1878, a law project passed which introduced inheritance taxation for lineal descendants. 
  - In 1893, liberal government introduced the first income taxation. 

- These precedents set the path for successive increases in rates based on the architecture set by the two preceding laws.

## Personal Impact on Politicians

```{r, out.width="0.7\\linewidth", include=TRUE, fig.align="center", fig.cap=c("Tax Rates on Income (Left) and Wealth (Right)"), echo=FALSE}
knitr::include_graphics("../figures/tax_rates.pdf")
```

# Method

## Empirical Model

- I firstly pool all laws, and model $V = \text{Pr}(p_i = 1)$ as a function of a politician's wealth and party, augmented by controls:

\[
V_{i, j} = \alpha + \beta \cdot \text{ihs } \text{Wealth}_{i,j} + \delta \cdot \text{Party}_{i} + \gamma \cdot \text{LawDum}_{j} + \eta \cdot \text{Controls}_{i,j} + \varepsilon_{i,j}
\]

- In other specifications, I separately estimate regressions for suffrage extensions and fiscal legislation.

- I use the following recursive relationship to estimate a politician's wealth at the time of voting on law $k$ as a function of their (deflated) wealth at death: 

\[
\text{Wealth}_{i,t+1} = \sum_{J} \text{AssetShare}_{i,j,t} \cdot \text{AssetReturn}_{i,j,[t,t+1]}
\]


## Endogeneity 

- Even after correcting for differential wealth returns, politicians' wealth could be endogenously determined, because particular voting behavior might be rewarded by interest groups, and other behavior is not (Fisman et al., 2014, Tahoun and van Lent, 2019)

- I use a dummy variable indicating whether a politician died within $x \in \{2,5\}$  years after having cast the vote on a particular law, and estimate the following model:

\begin{multline*}
V_{i, j} = \alpha + \beta_1 \cdot \text{ihs } \text{Wealth}_{i,j} + \beta_2 \cdot \text{Died within X years}_i + \\
\beta_3 \cdot \text{ihs } \text{Wealth x Died within X years }_{i,j} + \\
\beta_4 \cdot \text{Party}_i + \beta_5 \cdot \text{LawDum}_j + \gamma \cdot \text{Controls}_{i,j} + \varepsilon_{i,j}
\end{multline*}

## Endogeneity

- There might still be many reasons why the this approach does not isolate the effect of personal wealth on voting behavior. 
  - For example, politicians’ consumption and investment behavior might be correlated with their voting behavior: politicians who vote against might have consumed more of their income, reducing the observed correlation between personal wealth and  voting  behavior.  
  - More generally, unobserved politician fixed effects might be correlated with wealth. 
  
- I use instrumental variable estimation to address this. 
  - Two unrelated instruments: _Father Politician_ and _Expected Inheritance_

- Threats to identification imply that instrument might in itself suffer from endogeneity bias.
  - Politicians whose fathers were politically active could share a latent encompassing ideology (e.g. a penchant for increasing the size of the government)
  - They could therefore be inclined to vote in favor of laws that expand government
  - Robustness checks to make exclusion restriction plausible

# Results

## Descriptive Statistics

\input{../tables/descriptivestats_dissent.tex}

## OLS Estimates

- OLS Estimates of Wealth on the Propensity to Vote for Suffrage and Fiscal Legislation
  - Results show small but significant relationship

\begin{table}[!h]

\centering
\resizebox{\paperheight}{!}{
\begin{tabular}[t]{lccccc}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{3}{c}{Pooled} & \multicolumn{1}{c}{Suffrage} & \multicolumn{1}{c}{Fiscal} \\
\cmidrule(l{3pt}r{3pt}){2-4} \cmidrule(l{3pt}r{3pt}){5-5} \cmidrule(l{3pt}r{3pt}){6-6}
  & Model 1 & Model 2 & Model 3 & Model 4 & Model 5\\
\midrule
Personal Wealth &  & \num{-0.007}*** &  & \num{-0.008} & \num{-0.009}*\\
 &  & (\num{0.003}) &  & (\num{0.005}) & (\num{0.005})\\
Personal Wealth x Fiscal &  &  & \num{-0.007}** &  & \\
 &  &  & (\num{0.004}) &  \vphantom{1} & \\
Personal Wealth x Suffrage &  &  & \num{-0.007}* &  & \\
 &  &  & (\num{0.004}) &  & \\
Liberal & \num{0.385}*** & \num{0.433}*** & \num{0.433}*** & \num{0.383}*** & \num{0.373}***\\
 & (\num{0.027}) & (\num{0.033}) & (\num{0.033}) & (\num{0.071}) & (\num{0.062})\\
Socialist & \num{0.523}*** & \num{0.602}*** & \num{0.602}*** & \num{0.568}*** & \num{0.281}***\\
 & (\num{0.034}) & (\num{0.047}) & (\num{0.047}) & (\num{0.113}) & (\num{0.080})\\
\midrule
Law Fixed Effects & Yes & Yes & Yes & Yes & Yes\\
N & \num{1028} & \num{675} & \num{675} & \num{249} & \num{270}\\
Adj. R2 & \num{0.35} & \num{0.36} & \num{0.36} & \num{0.29} & \num{0.43}\\
\bottomrule
\multicolumn{6}{l}{\rule{0pt}{1em}Vote is defined as 1 if the politician is in favor of the reform, 0 otherwise.}\\
\multicolumn{6}{l}{\rule{0pt}{1em}The reference political allegiance is confessional. Personal Wealth is defined as ihs(Wealth at Time of Vote).}\\
\multicolumn{6}{l}{\rule{0pt}{1em}Heteroskedasticity-robust standard errors in parentheses. Results for lower house voting outcomes.}\\
\multicolumn{6}{l}{\rule{0pt}{1em}Control variables omitted from the table.}\\
\multicolumn{6}{l}{\rule{0pt}{1em}* p $<$ 0.1, ** p $<$ 0.05, *** p $<$ 0.01}\\
\end{tabular}}
\end{table}

## Endogeneity Test

\begin{table}

\caption{\label{tab:ols_endogeneity}Endogeneity Test for Suffrage Extension and Fiscal Legislation}
\centering
\resizebox{\paperheight}{!}{
\begin{tabular}[t]{lcccccc}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Pooled} & \multicolumn{2}{c}{Suffrage} & \multicolumn{2}{c}{Fiscal} \\
\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5} \cmidrule(l{3pt}r{3pt}){6-7}
  & Model 1 & Model 2 & Model 3 & Model 4 & Model 5 & Model 6\\
\midrule
Personal Wealth & \num{-0.008}* & \num{-0.010}** & \num{-0.012}* & \num{-0.013}* & \num{-0.009} & \num{-0.016}*\\
 & (\num{0.004}) & (\num{0.004}) & (\num{0.007}) & (\num{0.007}) & (\num{0.006}) & (\num{0.008})\\
Died W 2 Yrs & \num{0.028} & \num{0.033} & \num{0.032} & \num{0.023} & \num{0.031} & \num{-0.036}\\
 & (\num{0.063}) & (\num{0.069}) & (\num{0.109}) & (\num{0.107}) & (\num{0.095}) & (\num{0.107})\\
Personal Wealth x Died W 2 Yrs & \num{0.002} & \num{0.006} & \num{0.011} & \num{0.013} & \num{0.002} & \num{0.009}\\
 & (\num{0.006}) & (\num{0.006}) & (\num{0.009}) & (\num{0.009}) & (\num{0.008}) & (\num{0.009})\\
Liberal & \num{0.457}*** & \num{0.376}*** & \num{0.365}*** & \num{0.361}*** & \num{0.389}*** & \num{0.368}***\\
 & (\num{0.034}) & (\num{0.045}) & (\num{0.070}) & (\num{0.072}) & (\num{0.058}) & (\num{0.061})\\
Socialist & \num{0.496}*** & \num{0.381}*** & \num{0.612}*** & \num{0.600}*** & \num{0.327}*** & \num{0.277}***\\
 & (\num{0.043}) & (\num{0.060}) & (\num{0.106}) & (\num{0.111}) & (\num{0.072}) & (\num{0.086})\\
\midrule
Law Fixed Effects & Yes & Yes & Yes & Yes & Yes & Yes\\
N & \num{629} & \num{571} & \num{257} & \num{249} & \num{314} & \num{275}\\
Adj. R2 & \num{0.33} & \num{0.36} & \num{0.30} & \num{0.31} & \num{0.41} & \num{0.42}\\
\bottomrule
\multicolumn{7}{l}{\rule{0pt}{1em}Vote is defined as 1 if the politician is in favor of the reform, 0 otherwise.}\\
\multicolumn{7}{l}{\rule{0pt}{1em}The reference political allegiance is confessional. Personal Wealth is defined as ihs(Wealth at Time of Vote).}\\
\multicolumn{7}{l}{\rule{0pt}{1em}Heteroskedasticity-robust standard errors in parentheses. Results for lower house voting outcomes.}\\
\multicolumn{7}{l}{\rule{0pt}{1em}Control variables omitted from the table.}\\
\multicolumn{7}{l}{\rule{0pt}{1em}* p $<$ 0.1, ** p $<$ 0.05, *** p $<$ 0.01}\\
\end{tabular}}
\end{table}

## IV Estimates

## Robustness Checks




  
<!--
## Descriptive Statistics

- Preliminary evidence of the importance of wealth


## Descriptive Statistics 

- Variation in the degree of unanimity per party

\input{../tables/descriptive_stats_per_party_lh.tex}

## Descriptive Statistics

- Variation is lower in upper house

\input{../tables/descriptive_stats_per_party_uh.tex}

# Results

## Results - Baseline model

- This estimate addresses the influence of wealth  _within-party_.
- Plausible magnitude: a 1% increase in wealth decreases the probability of voting in favor with 1.8 percentage point.

## Baseline Results - Full controls (TK)

- Effect keeps existing after controlling for religion, region, constituent interests, political position


## Baseline Results - Full controls (EK)

- In the Upper house, the results are less robust to the inclusion of controls
  - Though the coefficient borders on significance and the point estimate is negative


## Alternative Explanations

- The effect of personal wealth on the likelihood to vote in favor of fiscal legislation seems to be robust to many controls. 
- But: politicians who vote against could have been rewarded by interest groups 
  - Wealth at death higher because of voting profile: coefficient might be _overestimated_
  
- Analysis of politicians who died shortly after last vote
  - No opportunity to amass rents
  
- IV using parental wealth
  - Parental wealth is a good proxy for wealth at the time of voting
  - Parental wealth correlated with politicians' wealth, but no direct effect on voting behavior.

## Alternative Explanations: Died Shortly After Vote

- Dying shortly after leaving office leaves little time to accumulate wealth from interest groups
  

## Alternative Explanations: Parental Wealth

- Wealth of politicians' parents proxies for politicians' wealth before voting
    - No direct influence on politicians' voting behavior (other than through politicians' wealth)

## IV Estimates (Preliminary)

\label{iv}

# Conclusion

## Conclusion

- I find a significant and persistent effect of personal interests on voting behavior of politicians.

  - The counterfactual implies that if a politician's wealth would increase by 20%, the effects of personal interests subsume party or ideological allegiance. 
  - The effect seems to be persistent in the lower house. 

- Politicians prioritize not only their ideology, but also their finances
  - But only when faced with imminent and real prospects of losing/winning
  - No convincing "general equilibrium" effects
  
- Confessional politicians have less party discipline

- Future: more observations in IV regression
  
# Thank you for your attention!

# Appendix: Fiscal reform

## How far did fiscal reforms reach?

\label{fiscal}

- After about 1870, majority of government revenues came from __taxation__. 


## Fiscal Reforms

- The share of tax revenue that came from real estate and wealth remained low
  - Steadily rising after the introduction of the 1893 income tax
  - Fiscal reforms encompassed (mildly) _progressive_ taxes
  - Lagged effects


## Fiscal Reforms

- Variation in plausible impact on politicians' personal wealth:

- Politicians' wealth would have certainly have been impacted by the Successiewet:
  - Tariffs up to 10 % (1878)
  - The poorest (net wealth < 1,000 guilders) were exempted
  - The 1911 revision implied a sharp increase in tariffs for lineal descendants

- The income tax was progressive, but the maximum tariff implied liability of only 3,2% of earned income (Fritschy, 2005)

- The revision in 1914 implied a maximum marginal tariff of 5%. (Own calculations, IB1914)

# Appendix: Data

## Voting Records

\label{votingrecords}

- Voting behavior in various important votes that lead to the acceptance or rejection of various far-reaching laws throughout the period 1870-1920.

- _Handelingen_, [statengeneraaldigitaal.nl](http://www.statengeneraaldigitaal.nl)
  
  - Find the dates on which the final roll call vote on a law took place

  - Data contain a _verbatim_ transcription of debates leading up to a vote
  
  - List of names of politicians who voted in favor, and against

## Politicians' Wealth

\label{memories}

- I use the _Memories van Successie_: probate inventories were administered by the Dutch tax administration for the purpose of levying inheritance taxes 
    - Universal from 1877 onwards
    - Publicly available until 1927

- Detailing the inventory of all assets and debts held by the deceased. 

- Detailed description also allows us to research the effect of **asset composition** on likelihood to vote in favor of fiscal legislation. 

## Controls

- Data regarding politicians' careers, district representation, ideological and party affiliation, and social origin from the [Politiek Documentatie Centrum](https://www.pdc.nl/)

- Electoral controls: turnout, nearest competitor, newspaper recommendation, etc. (for lower house members) from [this repository](http://resources.huygens.knaw.nl/verkiezingentweedekamer) by Ron De Jong

- Data on the frequency and location of strikes from the [IISG Labour Conflicts database](https://collab.iisg.nl/web/labourconflicts/datafiles)

- Data on demographic and economic characteristics of districts from a [database administered by Onno Boonstra](https://datasets.iisg.amsterdam/dataverse/HDNG?q=&types=files&sort=dateSort&order=desc&page=1)
    - Aggregated from the municipality to the district level (if applicable)


# Appendix: Spatial descriptive statistics

## Descriptive Statistics

# Appendix: Method

## Empirical Challenges

- Reverse causality: 
    - Politicians' voting behavior might be rewarded by interest groups. 
    - Politicians who vote against laws
    - Risk to __overestimate__ the effect of personal wealth on voting behavior
    
- Instrumental variable: parental wealth
    - Assumed to be strongly correlated with politicians' wealth before voting
    - No direct influence on politicians' voting behavior (other than through politicians' wealth)
    
## Empirical Model

- Baseline model (politician $i$, roll-call vote $k$):

\[
V_{i, k} = \alpha + \beta_1 \cdot \text{Wealth}_{i,k} + \beta_2 \cdot \text{Wealth}_{i,k} \cdot \text{LawDum}_k + \gamma \cdot \text{Controls}_{i,k} + u_{i,k}
\]

- In other words, I pool laws and estimate a linear probability model

- Calculate wealth at the time of voting from wealth at death:
\[
\text{Wealth}_{i,t+1} = \sum_{J} \text{AssetShare}_{i,j,t} \cdot \text{AssetReturn}_{i,j,[t,t+1]}
\]

- Since I know wealth at death, and the portfolio composition, calculate the (average) real returns on asset class $j$ and estimate the wealth at the time of voting.

- Historical returns provided by Jorda et al. (2019)


# Appendix: Additional results

## Robustness Checks: Other Laws




## Robustness Checks: Social Redistribution

\label{otherresults}

## Robustness Checks: Suffrage Expansion


## Robustness Checks: Wealth Composition


# The End

-->

