---
title: "Personal Wealth and Voting Behavior of Politicians"
author: "Bas Machielsen"
institute: "Utrecht University"
email: "a.h.machielsen@uu.nl"
date: "1/29/2021"
output: 
  beamer_presentation:
    includes:
      in_header: preamble.tex
classoption: "aspectratio=169"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

figures <- readRDS("../figures/figures_presentation.RDS")
```

# Motivation


## Motivation

- Many (Western) European countries underwent radical changes between roughly 1850 and 1920.

    - All of them experienced high economic growth and were rapidly industrializing.
    - Most of them transitioned to universal suffrage
    - Some of them established the groundwork of social security.

- Politicians can either obstruct or facilitate these changes

- What determines the behavior of politicians to do so?


## Motivation

- Common visions on what determines the voting behavior of politicians:
    
    - Electoral responsiveness: the principal-agent view of politics implies that well-monitoring politicians act in the interest of their principal, the electorate (Peltzman, 1982; Levitt, 1996)
    
    - Special cases of which are the threat of revolution (Aidt & Franck, 2019), or more broader constituent interests (Kalt & Zupan, 1984; Mian et al., 2010)
    
    - Ideology: politicians have an innate conviction or preference for a certain degree of redistribution.
    
- In this paper:

    - Does self-interest of politicians influence their voting behavior?
    
## Motivation

- I focus on several important laws that instigated __fiscal reforms__ in the Netherlands throughout the late 19th and early 20th centuries

- Excellent case to find out whether politicians prioritize their self-interest:

    - Fiscal reforms are very likely to affect politicians' personal finances. 
    - Various parliaments, laws with varying impact
    - Other high-profile laws to check robustness
    
- Implications for the role of politicians in (inclusive) economic development 

# Fiscal Reforms

## How far did fiscal reforms reach?

- After about 1870, majority of government revenues came from __taxation__. 

```{r echo=FALSE, out.width="250px", out.height="175px", fig.align='center'}
figures[[1]]
```

## Fiscal Reforms

- The share of tax revenue that came from real estate and wealth remained low
  - Steadily rising after the introduction of the 1893 income tax
  - Fiscal reforms encompassed (mildly) _progressive_ taxes
  - Lagged Effects

```{r echo = FALSE, out.width="250px", out.height="150px", fig.align='center'}
figures[[2]]
```

## Fiscal Reforms

- Variation in plausible impact on politicians' personal wealth:

- Politicians' wealth would have certainly have been impacted by the Successiewet:
  - Tariffs up to 10 % (1878)
  - The poorest (net wealth < 1,000 guilders) were exempted
  - The 1911 implied a sharp increase in tariffs for lineal descendants

- The income tax was progressive, but the maximum tariff implied liability of only 3,2% of earned income (Fritschy, 2005)

- The revision in 1914 implied a maximum marginal tariff of 5%. (Own calculations, IB1914)

# Data

## Voting Records

- My principal research question is about the influence of politicians' self-interest on their voting behavior in various important votes that lead to the acceptance or rejection of various far-reaching laws throughout the period 1870-1920.

- More precisely, I attempt to identify the influence of politicians' personal wealth on their voting behavior on laws that **potentially affect** politicians' personal wealth.

- By paying close attention to the theory, I attempt to incorporate as many insights in the behavior of politicians as possible.

- I try to measure both factors suggested by political historians, as well as factors suggested by economists and political scientists.

## Descriptive Statistics

- 

## Politicians' Wealth

- I use the _Memories van Successie_: probate inventories were administered by the Dutch tax administration for the purpose of levying inheritance taxes 
    - Universal from 1877 onwards
    - Publicly available until 1927

- Detailing the inventory of all assets and debts held by the deceased. 

- Detailed description also allows us to research the effect of **asset composition** on likelihood to vote in favor of fiscal legislation. 

## Controls

- Data regarding politicians' careers, district representation, ideological and party affiliation, and social origin from the [Politiek Documentatie Centrum](https://www.pdc.nl/)

- Electoral controls: turnout, nearest competitor, newspaper recommendation, etc. (for lower house members) from [this repository](http://resources.huygens.knaw.nl/verkiezingentweedekamer) by Ron De Jong

- Data on the frequency and location of strikes from the [IISG Labour Conflicts database](https://collab.iisg.nl/web/labourconflicts/datafiles)

- Data on demographic and economic characteristics of districts from a [database administered by Onno Boonstra](https://datasets.iisg.amsterdam/dataverse/HDNG?q=&types=files&sort=dateSort&order=desc&page=1)
    - Aggregated from the municipality to the district level (if applicable)


# Method

## Empirical Challenges

- Reverse causality: 
    - Politicians' voting behavior might be rewarded by interest groups. 
    - Politicians who vote against laws
    - Risk to __overestimate__ the effect of personal wealth on voting behavior
    
- Instrumental variable: parental wealth
    - Assumed to be strongly correlated with politicians' wealth before voting
    - No direct influence on politicians' voting behavior (other than through politicians' wealth)
    
## Empirical Model

- Baseline model (politician $i$, roll-call vote $k$):

\[
V_{i, k} = \alpha + \beta_1 \cdot \text{Wealth}_{i,k} + \beta_2 \cdot \text{Wealth}_{i,k} \cdot \text{LawDum}_k + \gamma \cdot \text{Controls}_{i,k} + u{i,k}
\]

- In other words, I pool laws and estimate a linear probability model

- Calculate wealth at the time of voting from wealth at death:
\[
\text{Wealth}_{i,t+1} = \sum_{J} \text{AssetShare}_{i,j,t} \cdot \text{AssetReturn}_{i,j,[t,t+1]}
\]

- Since I know wealth at death, and the portfolio composition, calculate the (average) real returns on asset class $j$ and estimate the wealth at the time of voting.

- Historical returns provided by Jorda et al. (2019)


# Results

## Baseline Results

## IV Estimates (Preliminary)

## Robustness Checks: Other Laws

[Effect on Electoral Expansion (0)]

[Effect on Social Intervention (0)]

## Robustness Checks: Wealth Composition

[With Real Estate Share etc. as Indep Vars]

# Conclusion

## Conclusion

