---
title: "Personal Wealth and Voting Behavior of Politicians"
author: "Bas Machielsen"
institute: "Utrecht University"
email: "a.h.machielsen@uu.nl"
date: "12 February 2021"
output: 
  beamer_presentation:
    includes:
      in_header: preamble.tex
classoption: "aspectratio=169"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(stargazer)
library(tidyverse)
library(ivreg)

figures <- readRDS("../figures/figures_presentation.RDS")
map1878 <- readRDS("../figures/presentation_map_1878.RDS")
map1893 <- readRDS("../figures/presentation_map_1893.RDS")

descr <- readRDS("../figures/descr.RDS")

first_regs <- readRDS("../figures/first_regs.RDS")

soc_rd_tk <- readRDS("../figures/social_redistribution_regressions.RDS")
elect_law_res_tk <- readRDS("../figures/electoral_law_regressions.RDS")

tk_fullctrls <- readRDS("../figures/tk_fullctrls.RDS")
ek_fullctrls <- readRDS("../figures/ek_fullctrls.RDS")

ivresults <- readRDS("../figures/ivresults.RDS")
harnas <- readRDS("../figures/harnas.RDS")
second_order <- readRDS("../figures/second_order_regs.RDS")
```

# Context

## PhD Thesis

- Overarching theme: Politicians and personal interests
  - Consequences for economic growth & development (Netherlands, 19th and early 20th centuries)

- Ch. 1: The Wealth of the Dutch Political Elite

- Ch. 2: Does personal wealth influence voting behavior? (This paper)

- Ch. 3: Being elected in national politics $\rightarrow$ becoming richer?

# Motivation

## Motivation

- Theoretical:
    - Principal-agent views of politics hold that politicians are agents on behalf of the electoral (the principals) (Duggan & Martinelli, 2017)
    - Politicians can decide to pursue their own interest (Mian et al., 2010)
    - One of the forms of pursuing self-interest is by increasing (non-decreasing) their wealth 

- Historical:
    - Historiography points at party discipline and ideology as most important factors in late 19th and early 20th century (e.g. Lijphart 1975, De Jong, 1999)
    - Personal interests overlooked, even though:
      - Small political elite
      - Uncompetitive electoral system

## Results - Overview

- Major finding: richer politicians are more likely to vote against fiscal legislation than _ceteris paribus_ poorer politicians

  - There is (likely) a direct effect of legislation on personal future wealth \hyperlink{fiscal}{\beamerbutton{Background}}

- The effects seem to be absent in legislation _not directly affecting_ politicians' personal wealth:

  - Suffrage expansion, other gov't intervention \hyperlink{otherresults}{\beamerbutton{Results}}
  
- Consistent with the historiography, ideology and party are major determinants of voting behavior

  - But the effect of personal wealth implies that a 20% increase/decrease would suffice to dominant the effect of party affiliation. 

## Approach

- Hypothesis: personal wealth influences voting behavior

- Voting behavior: voting records from lower & upper house \hyperlink{votingrecords}{\beamerbutton{Voting Data}}

  - Wealth: \textit{Memories van Successie} \hyperlink{memories}{\beamerbutton{Wealth Data}}
  - Control variables suggested by the literature
  
- Robustness checks: 

  - Take into account alternative explanations (interest-groups)
  - Compare with laws that do not directly affect personal finances
  
## Descriptive Statistics

- Preliminary evidence of the importance of wealth

```{r echo = FALSE}
descr[[1]] %>%
  knitr::kable(caption = "Eerste Kamer", 
               digits = 0)

```

## Descriptive Statistics

```{r echo = FALSE}
descr[[2]] %>%
  knitr::kable(caption = "Tweede Kamer",
               digits = 0)
```

# Results

## Results - Baseline model

- This estimate addresses the influence of wealth  _within-party_.
- Plausible magnitude: a 1% increase in wealth decreases the probability of voting in favor with 1.8 percentage point.

```{r echo = FALSE, results='asis'}
stargazer(first_regs,
          covariate.labels = c("Wealth"),
          dep.var.labels = "Vote",
          omit = c("law", "class"),
          omit.stat = c("adj.rsq", "ser","f"),
          add.lines = list(c("House", "Both", "Both", "TK", "EK"),
                           c("Controls", "None", "Party", "Party", "Party")
          ),
          notes.append = T,
          notes = c("Robust standard errors in parentheses"),
          header = F,
          font.size = "tiny",
          column.sep.width = "1pt",
          title = "Baseline regressions"
)

```

## Baseline Results - Full controls (TK)

- Effect keeps existing after controlling for religion, region, constituent interests, political position

```{r echo = FALSE, results = 'asis'}
stargazer(tk_fullctrls,
          covariate.labels = c("Wealth"),
          dep.var.labels = "Vote",
          omit = c("law", "class", "rk_pct", "agricul_share","strikes"),
          omit.stat = c("adj.rsq", "ser","f"),
          add.lines = list(c("House", "TK", "TK", "TK", "TK"),
                           c("Basic Controls", "Law+Party", "Law+Party", "Law+Party", "Law+Party"),
                           c("Other Controls", "None", "Strikes", "2 + Religion", "3 + Economy")
          ),
          notes.append = T,
          header = F,
          font.size = "tiny",
          column.sep.width = "1pt",
          title = "Estimates with various sets of controls"
          )
```

## Baseline Results - Full controls (EK)

- In the Upper house, the results are less robust to the inclusion of controls
  - Though the coefficient borders on significance and the point estimate is negative
  
```{r echo = FALSE, results='asis'}
stargazer(ek_fullctrls,
          covariate.labels = c("Wealth"),
          dep.var.labels = "Vote",
          omit = c("law", "class", "age_of_entrance", "age_of_vote", "tenure"),
          omit.stat = c("adj.rsq", "ser","f"),
          add.lines = list(c("House", "EK", "EK", "EK", "EK"),
                           c("Controls", "Party", "1 + Tenure", "2 + Age of Vote", "3 + Age of Entry")
          ),
          notes.append = T,
          notes = c("Robust standard errors in parentheses"),
          header = F,
          font.size = "tiny",
          column.sep.width = "1pt",
          title = "Estimates with various sets of controls")

```

## Alternative Explanations: Died Shortly After Vote

- Dying shortly after leaving office leaves little time to accumulate wealth from interest groups
  
```{r echo = FALSE, results='asis'}
stargazer(harnas,
          column.labels = c("<2 yr", "<2 yr", "<5 yr", "<5 yr"),
          covariate.labels = c("Wealth x No", "Wealth x Yes"),
          dep.var.labels = "Vote",
          omit = c("law", "class"),
          omit.stat = c("adj.rsq", "ser","f"),
          add.lines = list(c("House", "Both", "Both", "Both", "Both"),
                           c("Controls", "Party", "Party+Law",   "Party", "Party+Law")
                           ),
          notes.append = T,
          notes = c("Robust standard errors in parentheses"),
          header = F,
          font.size = "tiny",
          column.sep.width = "1pt",
          title = "Robustness analyses")

```

## Alternative Explanations: Parental Wealth

- Wealth of politicians' parents proxies for politicians' wealth before voting
    - Assumed to be strongly correlated with politicians' wealth
    - No direct influence on politicians' voting behavior (other than through politicians' wealth)
    
```{r echo = FALSE, warning = FALSE, out.width="160px", out.height="120px", fig.align='center'}
ivresults[[1]]
```

## IV Estimates (Preliminary)

\label{iv}

- No significance yet, but same magnitude as OLS estimates. 

```{r echo = FALSE, warning = FALSE, message = FALSE, results = 'asis'}
stargazer(ivresults[[2]], ivresults[[3]],
          dep.var.labels = c("Wealth", "Vote"),
          omit = c("class"),
          omit.stat = c("adj.rsq", "ser","f"),
          notes.append = T,
          header = F,
          font.size = "tiny",
          column.sep.width = "1pt",
          add.lines = list(c("Controls", "Party", "Party")),
          title = "IV estimates")
```

# Conclusion

## Conclusion

- I find a significant and persistent effect of personal interests on voting behavior of politicians.

  - The counterfactual implies that if you make a politician 20% richer, personal interests subsume party or ideological allegiance. 
  - The effect seems to be persistent in the **lower house**. 

- Politicians prioritize not only their ideology, but also their finances
  - But only when faced with imminent and real prospects of losing/winning
  - No convincing "general equilibrium" effects
  
- Confessional politicians have less party discipline

- Future: more observations in IV regression
  
# Thank you for your attention!

# Appendix: Fiscal reform

## How far did fiscal reforms reach?

\label{fiscal}

- After about 1870, majority of government revenues came from __taxation__. 

```{r echo=FALSE, out.width="250px", out.height="175px", fig.align='center'}
figures[[1]]
```

## Fiscal Reforms

- The share of tax revenue that came from real estate and wealth remained low
  - Steadily rising after the introduction of the 1893 income tax
  - Fiscal reforms encompassed (mildly) _progressive_ taxes
  - Lagged effects

```{r echo = FALSE, out.width="250px", out.height="150px", fig.align='center'}
figures[[2]]
```

## Fiscal Reforms

- Variation in plausible impact on politicians' personal wealth:

- Politicians' wealth would have certainly have been impacted by the Successiewet:
  - Tariffs up to 10 % (1878)
  - The poorest (net wealth < 1,000 guilders) were exempted
  - The 1911 revision implied a sharp increase in tariffs for lineal descendants

- The income tax was progressive, but the maximum tariff implied liability of only 3,2% of earned income (Fritschy, 2005)

- The revision in 1914 implied a maximum marginal tariff of 5%. (Own calculations, IB1914)

# Appendix: Data

## Voting Records

\label{votingrecords}

- Voting behavior in various important votes that lead to the acceptance or rejection of various far-reaching laws throughout the period 1870-1920.

- _Handelingen_, [statengeneraaldigitaal.nl](http://www.statengeneraaldigitaal.nl)
  
  - Find the dates on which the final roll call vote on a law took place

  - Data contain a _verbatim_ transcription of debates leading up to a vote
  
  - List of names of politicians who voted in favor, and against

## Politicians' Wealth

\label{memories}

- I use the _Memories van Successie_: probate inventories were administered by the Dutch tax administration for the purpose of levying inheritance taxes 
    - Universal from 1877 onwards
    - Publicly available until 1927

- Detailing the inventory of all assets and debts held by the deceased. 

- Detailed description also allows us to research the effect of **asset composition** on likelihood to vote in favor of fiscal legislation. 

## Controls

- Data regarding politicians' careers, district representation, ideological and party affiliation, and social origin from the [Politiek Documentatie Centrum](https://www.pdc.nl/)

- Electoral controls: turnout, nearest competitor, newspaper recommendation, etc. (for lower house members) from [this repository](http://resources.huygens.knaw.nl/verkiezingentweedekamer) by Ron De Jong

- Data on the frequency and location of strikes from the [IISG Labour Conflicts database](https://collab.iisg.nl/web/labourconflicts/datafiles)

- Data on demographic and economic characteristics of districts from a [database administered by Onno Boonstra](https://datasets.iisg.amsterdam/dataverse/HDNG?q=&types=files&sort=dateSort&order=desc&page=1)
    - Aggregated from the municipality to the district level (if applicable)


# Appendix: Spatial descriptive statistics

## Descriptive Statistics

- A correlation between region (religion) and voting behavior?

```{r echo = FALSE, out.width="300px", out.height="150px", fig.align='center'}
map1878
```

## Descriptive Statistics

```{r  echo = FALSE, out.width="300px", out.height="150px", fig.align='center'}
map1893
```

# Appendix: Method

## Empirical Challenges

- Reverse causality: 
    - Politicians' voting behavior might be rewarded by interest groups. 
    - Politicians who vote against laws
    - Risk to __overestimate__ the effect of personal wealth on voting behavior
    
- Instrumental variable: parental wealth
    - Assumed to be strongly correlated with politicians' wealth before voting
    - No direct influence on politicians' voting behavior (other than through politicians' wealth)
    
## Empirical Model

- Baseline model (politician $i$, roll-call vote $k$):

\[
V_{i, k} = \alpha + \beta_1 \cdot \text{Wealth}_{i,k} + \beta_2 \cdot \text{Wealth}_{i,k} \cdot \text{LawDum}_k + \gamma \cdot \text{Controls}_{i,k} + u_{i,k}
\]

- In other words, I pool laws and estimate a linear probability model

- Calculate wealth at the time of voting from wealth at death:
\[
\text{Wealth}_{i,t+1} = \sum_{J} \text{AssetShare}_{i,j,t} \cdot \text{AssetReturn}_{i,j,[t,t+1]}
\]

- Since I know wealth at death, and the portfolio composition, calculate the (average) real returns on asset class $j$ and estimate the wealth at the time of voting.

- Historical returns provided by Jorda et al. (2019)


# Appendix: Additional results

## Robustness Checks: Social Redistribution

\label{otherresults}

```{r echo = FALSE, results = 'asis'}

stargazer(soc_rd_tk,
          covariate.labels = c("Wealth"),
          dep.var.labels = "Vote",
          omit = c("law", "class", "strikes", "rk_pct", "agricul_share", "ncm"),
          omit.stat = c("adj.rsq", "ser","f"),
          add.lines = list(c("Controls", "Strikes", "1+Religion", "2+Economy", "3+Electoral"),
                           c("House", "TK", "TK", "TK", "TK")
          ),
          #          notes.append = T,
          #          notes = c("Robust standard errors in parentheses"),
          header = F,
                    font.size = "tiny",
          column.sep.width = "1pt",
          title = "Social Redistribution")
```

## Robustness Checks: Suffrage Expansion

```{r echo = FALSE, results = 'asis'}

stargazer(elect_law_res_tk,
          covariate.labels = c("Wealth"),
          dep.var.labels = "Vote",
          omit = c("law", "class", "strikes", "rk_pct", "agricul_share", "ncm"),
          omit.stat = c("adj.rsq", "ser","f"),
          add.lines = list(c("Controls", "Strikes", "1+Religion", "2+Economy", "3+Electoral"),
                           c("House", "TK", "TK", "TK", "TK")
          ),
          #          notes.append = T,
          #          notes = c("Robust standard errors in parentheses"),
          header = F,
                    font.size = "tiny",
          column.sep.width = "1pt",
          title = "Suffrage Extension")

```

## Robustness Checks: Wealth Composition

```{r echo = FALSE, results='asis'}
stargazer(second_order,
          covariate.labels = c("Real Estate", "Foreign Assets", "Shares"),
          dep.var.labels = "Vote",
          omit = c("law", "class"),
          omit.stat = c("adj.rsq", "ser","f"),
          add.lines = list(c("House", "Both", "Both", "Both", "Both", "Both", "Both"),
                           c("Controls", "Party", "Party+Law", 
                             "Party", "Party+Law",
                             "Party", "Party+Law")
          ),
          notes.append = T,
          notes = c("Robust standard errors in parentheses"),
          header = F,
          font.size = "tiny",
          column.sep.width = "1pt",
          title = "Second-order effects"
)

```

# The End

